Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-03-19 paper/paper2.tex
Typo fix
minor

%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
\documentclass[journal=jpcbfk, layout=twocolumn, manuscript=article]{achemso}
\setkeys{acs}{articletitle = true}
\usepackage[utf8]{inputenc}

@@

\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
\author{Christian R. Schwantes}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Robert T. McGibbon}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Vijay S. Pande}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
\email{pande@stanford.edu}

@@

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. Here, we show that by consideration of the states' conditional distribution over conformations, the state space can be evaluated and optimized using the Markov state model's likelihood function and standard statistical model selection tools. Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
\end{abstract}

\section{Introduction}
Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic insight into conformational dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}. In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}. Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indicator functions. In particular, when states contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably. While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.

@@

\section{Prior Work}
%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases. The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.

@@

Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state. That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume. This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.

>>>>

%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
\documentclass[journal=jpcbfk, layout=traditional, manuscript=article]{achemso}
\setkeys{acs}{articletitle = true}
\usepackage[utf8]{inputenc}

@@

\title{Statistical Model Selection for Markov Models of Biomolecular Dynamics}
\author{Christian R. Schwantes}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Robert T. McGibbon}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Vijay S. Pande}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
\email{pande@stanford.edu}
\keywords{MSM, molecular dynamics, AIC, BIC, Bayes factor}

@@

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly. Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
\end{abstract}

\section{Introduction}
Proteins are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic insight into conformational dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles with relative ease\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}. A central challenge in the field of molecular simulations is now often the kinetic analysis of simulation datasets. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}. High-resolution  models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states which partition the phase space. Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

Some explanation of notation...
\begin{align}
s(t) &= \sigma(x(t)), \\
s(t+\tau) &\sim \mathbf{T}_{s(t), k} \sigma_k.
\end{align}

A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}. Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space does (such as the discrete indicator functions used by MSMs) generally experiences non-Markovian dynamics. In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel\cite{ZM}. Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system. In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large. While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.

@@

\section{Prior Work}
%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases. The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.

@@

Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state. That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume. This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the state.
